$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-02 paper/paper.tex
Change Carrier to Usage

% Types
\newcommand{\Set}{\type{SET}}
\newcommand{\reduce}[1]{\; \type{\longrightarrow}_{#1} \;}
\newcommand{\types}[4]{#1 \; \type{\propto} \; #2 \; \type{\vdash} \; #3 \; \type{\boxtimes} \; #4}
\newcommand{\contains}[6]{#1 \; \type{\propto} \; #2 \; \type{\ni}_{#3} \; #4 \; \type{\propto} \; #5 \; \type{\boxtimes} \; #6}
\newcommand{\containsusage}[4]{#1 \; \type{\ni}_{#2} \; #3 \; \type{\boxtimes} \; #4}
\newcommand{\Var}{\type{VAR}}
\newcommand{\Process}{\type{PROCESS}}
\newcommand{\Unused}{\type{UNUSED}}
\newcommand{\PreCtx}{\type{PRECTX}}
\newcommand{\Ctx}{\type{CTX}}
\newcommand{\Type}{\type{TYPE}\;}
\newcommand{\Idx}{\type{IDX}\;}
\newcommand{\Idxs}{\type{IDXS}}
\newcommand{\Carrier}{\type{CARRIER}}
\newcommand{\N}{\type{\mathbb{N}}}
\newcommand{\Channel}{\type{CHANNEL}}
\newcommand{\Rec}{\type{REC}}
\newcommand{\Quantifier}{\type{QUANTIFIER}}
\newcommand{\eq}[1]{\; \type{=}_{#1} \;}
\newcommand{\eqeq}{\; \type{\equiv} \;}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{$\pi$ with leftovers: \\ a mechanisation in Agda}
\titlerunning{$\pi$ with leftovers: all mechanisation in Agda}
\author{Uma Zalakain}{University of Glasgow, Scotland}
       {u.zalakain.1@research.gla.ac.uk}{https://orcid.org/0000-0002-3268-9338}{}
\author{Ornela Dardha}{University of Glasgow, Scotland}
       {ornela.dardha@glasgow.ac.uk}{https://orcid.org/0000-0001-9927-7875}{}
\authorrunning{U. Zalakain and O. Dardha}
\Copyright{Uma Zalakain and Ornela Dardha}
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003753.10003761.10003764</concept_id>
<concept_desc>Theory of computation~Process calculi</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}
\ccsdesc[300]{Theory of computation~Process calculi}
\keywords{pi-calculus, linear types, leftover typing, concurrency, mechanization, Agda}
\supplement{\url{https://github/umazalakain/typing-linear-pi}}
\acknowledgements{We want to thank Erika Kreuter, Wen Kokke, James Wood, Guillaume Allais, Bob Atkey, and Conor McBride for their thoughts, patience, work, education and camaraderie.}
\funding{Supported by the UK EPSRC grant EP/K034413/1, ``From Data Types to Session Types: A Basis for Concurrency and Distribution'' (ABCD), and by the EU HORIZON 2020 MSCA RISE project 778233
  ``Behavioural Application Program Interfaces'' (BehAPI).}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{note}
  We will often work with pairs of input and output multiplicities.
  We use the notation $\Carrier^2$ to stand for a $\Carrier \times \Carrier$ tuple of input and output multiplicities, respectively.
  We use $\op{x}{y}{z}$ to stand for a monoidal operation on pairs of multiplicities -- where the algebraic laws are lifted elementwise.
  Henceworth, we use $\lz$ to denote the pair $(0 \comma 0)$, $\li$ for the pair $(1 \comma 0)$, $\lo$ for $(0 \comma 1)$ and $\lio$ for $(1 \comma 1)$, borrowing from the linear \picalc{} notation \cite{KPT96,Sangio01}.
\end{note}

@@

We keep typing contexts and usage contexts separate.
The former are preserved throughout a typing derivation; the latter are transformed as a result of input, output, and context splits.

@@

    \inferrule
    {t : \Type \\ \stacked{idx : \Idx \\\\ x : \Carrier_{idx}^2}}
    {\channel{t}{x} : \Type}
  \end{mathpar}

@@

    \inferrule
        {\Gamma : \Ctx_{idxs} \\ x : \Carrier_{idx} ^2}
        {\Gamma \comma x : \Ctx_{idxs \comma idx}}
  \end{mathpar}
  \caption{Length-indexed context of carrier indices with a context of multiplicities on top.}
  \label{fig:ctx}
  \end{figure}
\end{nidefinition}

@@

    \datatype{
      \gamma : \PreCtx_n \\
      \stacked{
        idxs : \Idxs_n \\\\
        \Gamma : \Ctx_{idxs}} \\
      i : \Var_n \\
      t : \Type \\
      \stacked{
        idx : \Idx \\\\
        y : \Carrier_{idx}^2} \\
      \Delta : \Ctx_{idxs}}
    {\contains{\gamma}{\Gamma}{i}{t}{y}{\Delta} : \Set}
    \; \textsc{VarRef}

@@

    \inferrule
    {t : \Type \\ x : \Carrier_{idx}^2 \\ y : \Carrier_{idx'} \\\\
     cont : \types{\gamma \comma \channel{t}{x}}{\Gamma \comma (y \comma y) }{P}{\Delta \comma \lz}}
    {\constr{chan} \; t \; x \; y \; cont : \types{\gamma}{\Gamma}{\new P}{\Delta}}

@@

Framing captures the idea that the well-typedness of a process only depends on the resources consumed by it.
These unused resources are not introduced as new variables -- the typing context remains unaltered -- but rather as extra multiplicities that are carried from the input context to the output context.

\begin{nitheorem}
  \label{thm:framing}
  Let $P$ be a well-typed process from context $\Gamma_l$ to $\Xi_l$.
  Then, per \autoref{lm:types-op}, there exists some $\Delta$ such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Take arbitrary contexts $\Gamma_r$ and $\Xi_r$ such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

Weakening states that adding new variables to the typing and usage contexts preserves the well-typedness of a process.
The multiplicities of these new variables will be carried from the input context to the output context.
Usually, weakening is presented through order preserving embeddings.
Order preserving embeddings model a series of insertions into a context.
In this paper, we use weakening to prove subject congruence.
To do so, we need of a single insertion into a context, and thus we will limit our theorem statement to that, but there is no loss of generality.

\begin{nitheorem}
  \label{thm:weakening}
  Let $P$ be a well-typed process such that $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let us insert an arbitrary type $t$ at position $i$ in the typing context and an arbitrary multiplicity $x$ at position $i$ into both the input and output contexts.
  Let us reflect this on the process $P$ by incrementing every index greater than or equal to $i$.
  The resulting process will be well-typed so that $\types{insert_i \; t \; \gamma}{insert_i \; x \; \Gamma}{lift_i \; P}{insert_i \; x \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

>>>>

% Types
\newcommand{\Set}{\type{SET}}
\newcommand{\reduce}[1]{\; \type{\longrightarrow}_{#1} \;}
\newcommand{\types}[4]{#1 \; \type{\propto} \; #2 \; \type{\vdash} \; #3 \; \type{\boxtimes} \; #4}
\newcommand{\contains}[6]{#1 \; \type{\propto} \; #2 \; \type{\ni}_{#3} \; #4 \; \type{\propto} \; #5 \; \type{\boxtimes} \; #6}
\newcommand{\containsusage}[4]{#1 \; \type{\ni}_{#2} \; #3 \; \type{\boxtimes} \; #4}
\newcommand{\Var}{\type{VAR}}
\newcommand{\Process}{\type{PROCESS}}
\newcommand{\Unused}{\type{UNUSED}}
\newcommand{\PreCtx}{\type{PRECTX}}
\newcommand{\Ctx}{\type{CTX}}
\newcommand{\Type}{\type{TYPE}\;}
\newcommand{\Idx}{\type{IDX}\;}
\newcommand{\Idxs}{\type{IDXS}}
\newcommand{\Usage}{\type{USAGE}}
\newcommand{\N}{\type{\mathbb{N}}}
\newcommand{\Channel}{\type{CHANNEL}}
\newcommand{\Rec}{\type{REC}}
\newcommand{\Quantifier}{\type{QUANTIFIER}}
\newcommand{\eq}[1]{\; \type{=}_{#1} \;}
\newcommand{\eqeq}{\; \type{\equiv} \;}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{$\pi$ with leftovers: \\ a mechanisation in Agda}
\titlerunning{$\pi$ with leftovers: all mechanisation in Agda}
\author{Uma Zalakain}{University of Glasgow, Scotland}
       {u.zalakain.1@research.gla.ac.uk}{https://orcid.org/0000-0002-3268-9338}{}
\author{Ornela Dardha}{University of Glasgow, Scotland}
       {ornela.dardha@glasgow.ac.uk}{https://orcid.org/0000-0001-9927-7875}{}
\authorrunning{U. Zalakain and O. Dardha}
\Copyright{Uma Zalakain and Ornela Dardha}
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003753.10003761.10003764</concept_id>
<concept_desc>Theory of computation~Process calculi</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}
\ccsdesc[300]{Theory of computation~Process calculi}
\keywords{pi-calculus, linear types, leftover typing, concurrency, mechanization, Agda}
\supplement{\url{https://github/umazalakain/typing-linear-pi}}
\acknowledgements{We want to thank Erika Kreuter, Wen Kokke, James Wood, Guillaume Allais, Bob Atkey, and Conor McBride for their thoughts, patience, work, education and camaraderie.}
\funding{Supported by the UK EPSRC grant EP/K034413/1, ``From Data Types to Session Types: A Basis for Concurrency and Distribution'' (ABCD), and by the EU HORIZON 2020 MSCA RISE project 778233
  ``Behavioural Application Program Interfaces'' (BehAPI).}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{note}
  We will often work with pairs of input and output multiplicities.
  We use the notation $\Usage^2$ to stand for a $\Usage \times \Usage$ tuple of input and output multiplicities, respectively.
  We use $\op{x}{y}{z}$ to stand for a monoidal operation on pairs of multiplicities -- where the algebraic laws are lifted elementwise.
  Henceworth, we use $\lz$ to denote the pair $(0 \comma 0)$, $\li$ for the pair $(1 \comma 0)$, $\lo$ for $(0 \comma 1)$ and $\lio$ for $(1 \comma 1)$, borrowing from the linear \picalc{} notation \cite{KPT96,Sangio01}.
\end{note}

@@

We keep typing contexts ($\PreCtx$) and usage contexts ($\Ctx$) separate.
The former are preserved throughout a typing derivation; the latter are transformed as a result of input, output, and context splits.

@@

    \inferrule
    {t : \Type \\ \stacked{idx : \Idx \\\\ x : \Usage_{idx}^2}}
    {\channel{t}{x} : \Type}
  \end{mathpar}

@@

    \inferrule
        {\Gamma : \Ctx_{idxs} \\ x : \Usage_{idx} ^2}
        {\Gamma \comma x : \Ctx_{idxs \comma idx}}
  \end{mathpar}
  \caption{Length-indexed context of carrier indices with a context of multiplicities on top.}
  \label{fig:ctx}
  \end{figure}
\end{nidefinition}

@@

    \datatype{
      \gamma : \PreCtx_n \\
      \stacked{
        idxs : \Idxs_n \\\\
        \Gamma : \Ctx_{idxs}} \\
      i : \Var_n \\
      t : \Type \\
      \stacked{
        idx : \Idx \\\\
        y : \Usage_{idx}^2} \\
      \Delta : \Ctx_{idxs}}
    {\contains{\gamma}{\Gamma}{i}{t}{y}{\Delta} : \Set}
    \; \textsc{VarRef}

@@

    \inferrule
    {t : \Type \\ x : \Usage_{idx}^2 \\ y : \Usage_{idx'} \\\\
     cont : \types{\gamma \comma \channel{t}{x}}{\Gamma \comma (y \comma y) }{P}{\Delta \comma \lz}}
    {\constr{chan} \; t \; x \; y \; cont : \types{\gamma}{\Gamma}{\new P}{\Delta}}

@@

Framing captures the idea that the well-typedness of a process is independent of its leftover resources.

\begin{nitheorem}
  \label{thm:framing}
  Let $P$ be a well-typed process $\types{\gamma}{\Gamma_l}{P}{\Xi_l}$.
  Then, per \autoref{lm:types-op}, there exists some $\Delta$ such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Take arbitrary contexts $\Gamma_r$ and $\Xi_r$ such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

Weakening states that adding new variables to the typing and input and output usage contexts preserves the well-typedness of a process.
The usage annotations of the new variables will be left untouched as leftovers.

\begin{nitheorem}
  \label{thm:weakening}
  Let $P$ be a well-typed process $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Insert an arbitrary type $t$ at position $i$ in the typing context, and an arbitrary usage annotation $x$ at position $i$ into both the input and output contexts.
  Increment every referrence to a variable greater than or equal to $i$ in process $P$.
  The resulting process will be well-typed so that $\types{insert_i \; t \; \gamma}{insert_i \; x \; \Gamma}{lift_i \; P}{insert_i \; x \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}
